General limits

### 2017-08-27 papers/talk-7/paper-7.tex
limit construction, colimit definition

\maketitle
\chapter{Introduction}
We have seen three special structures in categories.\\[1em]
\begin{minipage}{.35\linewidth}
  \begin{tikzcd}
    & V \arrow[bend right=40]{ddl} \arrow[bend left=40]{ddr} \arrow[dashed]{d} &\\
    & U \arrow{dl} \arrow{dr} &\\
    \bullet &&\bullet
  \end{tikzcd}
\end{minipage}%
\begin{minipage}{.35\linewidth}
  \begin{tikzcd}
    V
    \arrow[bend left=40]{drr}
    \arrow[bend right=40]{ddr}
    \arrow[dashed,swap]{dr}
    && \\
    &U
    \arrow{r}
    \arrow{d}
    &\bullet \arrow{d} \\
    & \bullet \arrow{r} & \bullet
  \end{tikzcd}
\end{minipage}%
\begin{minipage}{.3\linewidth}
  \begin{tikzcd}
    V \arrow[dashed]{dd} \arrow{dr}&& \\
    & \bullet \arrow[shift left]{r} \arrow[shift right]{r} & \bullet \\\
    U \arrow{ur} &&
  \end{tikzcd}
\end{minipage}\\[2em]
Now we generalize them...
\chapter{Definition}

In category theory we take limits over diagrams.
We begin by formally defining what a diagram is.

@@

The functor $D$ yields
a part of the category $\mathscr C$ that essentially looks like $\bm I$.
We also say $D$ \textbf{identifies} diagrams of shape $\bm I$,
which is why we use $\bm I$ to denote the shape.
Especially with diagrams of such a small shape
there might be many parts in $\mathscr C$ that have the shape of $\bm I$.
We therefore get all different diagrams of that shape by considering
all different functors $D$.
Diagrams do not need to be small though. Their shape can also be infinite.\\
We are typically interested in diagrams of a specific category,
$\mathscr C$ in the example. Therefore we usually leave out $\bm I$ and $D$,
and we just draw the identified part of $\mathscr C$ instead.

\begin{definition}
  Let $D \from \bm{I} \to \mathscr{C}$ be a diagram.
  A \textbf{cone} $U$ over $D$ is an object $U \in \mathscr{C}$
  with maps $f_I, \forall\ I \in \bm{I}$, such that
  $\forall\ I \overset{m}{\to} J \in \bm{I}$, the following triangle commutes:
  \[
    \begin{tikzcd}
      & U \arrow[dl, swap, "f_I"] \arrow{dr}{f_J}&\\
      D_I \arrow[rr, swap, "D(m)"] & & D_J\\
    \end{tikzcd}
  \]
  That is $f_J = D(m) \circ f_I$.\\
  For a cone we write $(U \overset{f_I}{\to} D(I))_{I\in \bm{I}}$.
  We call $U$ the \textbf{vertex}, and $f_I$ the \textbf{projection maps} of the cone.
\end{definition}
\begin{remark}
  When $U$ is the vertex of a cone, then we sometimes use just $U$ to refer to the whole cone
  for better readability.
\end{remark}
%% TODO: Notation DI, D(I), D_I ??
\begin{example}
  A cone over the diagram of
  \hyperref[diagramexample]{Example \ref*{diagramexample}} looks like this.
  \[
    \begin{tikzcd}
      U \arrow[bend right=30]{ddr}{f_I} \arrow[bend left=30]{drr}{f_k} \arrow[bend left=20, swap]{ddrr}{f_J}&   &\\
      & & D_K \arrow{d}{D_n} \\
      & D_I \arrow{r}{D_m} & D_J
    \end{tikzcd}
  \]
  The commuting condition says that $D_m \circ f_I = f_J$ and $D_n \circ f_K = f_J$.
\end{example}

@@

\begin{definition}
  A cone $(U \overset{f_I}{\to} D(I))_{I\in \bm{I}}$ is universal, if
  for all other cones $(V \overset{g_I}{\to} D(I))_{I\in \bm{I}}$ there is a unique
  map $h \from V \to U$, such that for all $I \in \bm{I}$
  the following triangles commute.
  \[
    \begin{tikzcd}
      V \arrow[dr, swap, "g_I"] \arrow{r}{h} & U \arrow{d}{f_I} \\
      & D_I
    \end{tikzcd}
  \]
  That is $g_i = f_i \circ h$.
\end{definition}

In other words, any cone over $D$ has maps into $D$, and a cone $U$ is universal if
for any other cone $V$, you can first go to $U$, that is the unique morphism.
From $U$ you can go to $D_I$, and the commuting condition says that this is equal
to the morphism that immediately takes you from $V$ to $D_I$.

\begin{example}
    \[
    \begin{tikzcd}
      {\color{red}V}
      \arrow[bend right=40,color=red]{dddrr}{g_I}
      \arrow[bend left=40, color=red]{ddrrr}{g_k}
      \arrow[bend left=30, color=red, swap]{dddrrr}{g_J}
      \arrow[dashed,swap]{dr}{h}
      & & & \\
      & {\color{blue}U}
      \arrow[bend right=30,color=blue]{ddr}{f_I}
      \arrow[bend left=30, color=blue]{drr}{f_k}
      \arrow[bend left=20, color=blue, swap]{ddrr}{f_J}&   &\\
      & & & D_K \arrow{d}{D_n} \\
      & & D_I \arrow{r}{D_m} & D_J
    \end{tikzcd}
  \]
  The universal cone {\color{blue}U}
  is drawn in blue and the other cone {\color{red}V} is drawn in red.
  Both cones obviously have to satisfy the commuting condition for cones.
  Additionally the commuting condition for universal cones states that
  $g_I = f_I \circ h$, $g_J = f_J \circ h$, and $g_K = f_k \circ h$.
\end{example}

\begin{definition}
  A limit is a universal cone over a diagram.
\end{definition}

@@

%% TODO: terminal object, pullback, equalizer, product.

\chapter{Building new limits}
It is possible to build new limits...
%% TODO: example
%% TODO: general limits
%% TODO: possibly other construction mentioned in awodey.

\chapter{Colimits}
Colimits are the dual of limits.
\begin{definition}
  A \textbf{colimit} is a universal co-cone over a diagram.
\end{definition}
%% TODO: complete definition

>>>>

%% TODO: consistent notation, DI D(I) D_I, also for maps D_m Dm

\maketitle
\chapter{Introduction}
We have seen three special structures in categories.\\[1em]
\begin{minipage}{.35\linewidth}
  \begin{tikzcd}
    & V \arrow[bend right=40]{ddl} \arrow[bend left=40]{ddr} \arrow[dashed]{d} &\\
    & U \arrow{dl} \arrow{dr} &\\
    \bullet &&\bullet
  \end{tikzcd}
\end{minipage}%
\begin{minipage}{.35\linewidth}
  \begin{tikzcd}
    V
    \arrow[bend left=40]{drr}
    \arrow[bend right=40]{ddr}
    \arrow[dashed,swap]{dr}
    && \\
    &U
    \arrow{r}
    \arrow{d}
    &\bullet \arrow{d} \\
    & \bullet \arrow{r} & \bullet
  \end{tikzcd}
\end{minipage}%
\begin{minipage}{.3\linewidth}
  \begin{tikzcd}
    V \arrow[dashed]{dd} \arrow{dr}&& \\
    & \bullet \arrow[shift left]{r} \arrow[shift right]{r} & \bullet \\\
    U \arrow{ur} &&
  \end{tikzcd}
\end{minipage}\\[2em]
These three examples follow a pattern.
In each example there is an object $U$ with some arrows. This object $U$
has a universal property which states that any other object $V$ which has arrows
similar to $U$
factors uniquely through $U$, that is, there is a unique morphism $V \to U$,
such that the resulting triangles commute.\\
Products, pullbacks, and equalizers are instances of a more general concept called
limit which expresses the common pattern between these examples.
%% TODO: possibly give further applications of limits too
%% TODO: maybe some more intuition what limits are about

\chapter{Definition}
In category theory we take limits over diagrams.
We begin by formally defining what a diagram is.

@@

The functor $D$ yields
a collection of objects and morphisms in the category $\mathscr C$
that essentially looks like $\bm I$.
We also say $D$ \textbf{identifies} diagrams of shape $\bm I$.
Especially with diagrams of such a small shape
there might be many parts in $\mathscr C$ that have the shape of $\bm I$.
We therefore get all different diagrams of that shape by considering
all different functors $D$.
Diagrams do not need to be small though. Their shape can also be infinite.\\
We are typically interested in diagrams of a specific category,
$\mathscr C$ in the example. Therefore we usually leave out $\bm I$ and $D$,
and we just draw the identified part of $\mathscr C$ instead.

\begin{definition}
  Let $D \from \bm{I} \to \mathscr{C}$ be a diagram.
  A \textbf{cone} $U$ over $D$ is an object $U \in \mathscr{C}$
  with maps $U \overset{f_I}{rightarrow} D_I, \forall\ I \in \bm{I}$, such that
  $\forall\ I \overset{m}{\to} J \in \bm{I}$, the following triangle commutes:
  \[
    \begin{tikzcd}
      & U \arrow[dl, swap, "f_I"] \arrow{dr}{f_J}&\\
      D_I \arrow[rr, swap, "D(m)"] & & D_J\\
    \end{tikzcd}
  \]
  That is $f_J = D(m) \circ f_I$.\\
  For a cone we write $(U \overset{f_I}{\to} D(I))_{I\in \bm{I}}$.
  We call $U$ the \textbf{vertex}, and $f_I$ the \textbf{projection maps} of the cone.
\end{definition}
\begin{remark}
  When $U$ is the vertex of a cone, then we sometimes use just $U$ to refer to the whole cone
  for better readability.
\end{remark}
%% TODO: Notation DI, D(I), D_I ??
\begin{example}
  A cone over the diagram of
  \hyperref[diagramexample]{Example \ref*{diagramexample}} looks like this.
  \[
    \begin{tikzcd}
      U \arrow[bend right=30]{ddr}{f_I} \arrow[bend left=30]{drr}{f_k} \arrow[bend left=20, swap]{ddrr}{f_J}&   &\\
      & & D_K \arrow{d}{D_n} \\
      & D_I \arrow{r}{D_m} & D_J
    \end{tikzcd}
  \]
  The commuting condition says that $D_m \circ f_I = f_J$ and $D_n \circ f_K = f_J$.
\end{example}

@@

\begin{definition}
  A cone $(U \overset{f_I}{\to} D(I))_{I\in \bm{I}}$ is \textbf{universal}, if
  for all other cones $(V \overset{g_I}{\to} D(I))_{I\in \bm{I}}$ there is a unique
  map $h \from V \to U$, such that for all $I \in \bm{I}$
  the following triangles commute.
  \[
    \begin{tikzcd}
      V \arrow[dr, swap, "g_I"] \arrow{r}{h} & U \arrow{d}{f_I} \\
      & D_I
    \end{tikzcd}
  \]
  That is $g_i = f_i \circ h$.
\end{definition}

In other words, any cone over $D$ has maps into $D$, and a cone $U$ is universal if
any other cone $V$ has a unique morphism to $U$.
The commuting condition states that for any cone the projection maps $g_I$ are equivalent
to the map that is obtained by composing the unique limit map $h$ with the
projection maps of the limit $f_I$.

\begin{example}
    \[
    \begin{tikzcd}
      {\color{red}V}
      \arrow[bend right=40,color=red]{dddrr}{g_I}
      \arrow[bend left=40, color=red]{ddrrr}{g_k}
      \arrow[bend left=30, color=red, swap]{dddrrr}{g_J}
      \arrow[dashed,swap]{dr}{h}
      & & & \\
      & {\color{blue}U}
      \arrow[bend right=30,color=blue]{ddr}{f_I}
      \arrow[bend left=30, color=blue]{drr}{f_k}
      \arrow[bend left=20, color=blue, swap]{ddrr}{f_J}&   &\\
      & & & D_K \arrow{d}{D_n} \\
      & & D_I \arrow{r}{D_m} & D_J
    \end{tikzcd}
  \]
  The universal cone {\color{blue}U}
  is drawn in blue and the other cone {\color{red}V} is drawn in red.
  Both cones obviously have to satisfy the commuting condition for cones.
  Additionally the commuting condition for universal cones states that
  $g_I = f_I \circ h$, $g_J = f_J \circ h$, and $g_K = f_k \circ h$.
  Therefore this diagram commutes in its entirety.
\end{example}

\begin{definition}
  A \textbf{limit} is a universal cone over a diagram.
\end{definition}

@@

\begin{example}
  Consider a diagram of the following shape
  \begin{tikzcd}
    \bullet &\bullet
  \end{tikzcd},
  which is just two objects without non-trivial arrows.
  A limit over this diagram looks like this.
  \[
    \begin{tikzcd}
      & {\color{red} V}
      \arrow[bend right=40,color=red, swap]{ddl}{g_I}
      \arrow[bend left=40,color=red]{ddr}{g_J}
      \arrow[dashed]{d}{h} &\\
      & {\color{blue} U}
      \arrow[color=blue, swap]{dl}{f_I}
      \arrow[color=blue]{dr}{f_J} &\\
      I && J
    \end{tikzcd}
  \]
  This is the ordinary binary product.
  The commuting condition for limits states that
  $f_I \circ h = g_I$, and $f_J \circ h = g_J$, which
  is exactly the commuting condition for products.
  We have therefore verified that the product is indeed an instance of the general limit definition.
  Using discrete categories with more than two objects we get products of arbitrary size.
  \[
    \begin{tikzcd}
      && {\color{red} V}
      \arrow[bend right=40,color=red]{ddll}
      \arrow[bend right=40,color=red]{ddl}
      \arrow[bend left=40,color=red]{ddrr}
      \arrow[bend left=40,color=red]{ddr}
      \arrow[dashed]{d} &&\\
      && {\color{blue} U}
      \arrow[color=blue]{dl}
      \arrow[color=blue]{dll}
      \arrow[color=blue]{dr}
      \arrow[color=blue]{drr} &&\\
      \bullet & \bullet & \ldots & \bullet &\bullet
    \end{tikzcd}
  \]  
\end{example}

\begin{example}
  If we take the empty category as the shape for a diagram, then a cone
  over that diagram is just a single object $U$ that has no other required arrows.
  Then the universal property for limits states that every other cone, i.e. an object $V$,
  has a unique map $V \to U$. A limit over the empty diagram is therefore just a terminal object,
  and terminal objects are therefore another instance of the general limit definition.
\end{example}

\begin{example}
  Consider the following diagram.
  \begin{tikzcd}
    I \arrow{r}{m} & J & K \arrow[swap]{l}{n}
  \end{tikzcd}
  A cone over this diagram looks like this:
  \[
    \begin{tikzcd}
      U
      \arrow[]{r}{\ \ \ f_K}
      \arrow[swap]{rd}{\ f_J}
      \arrow[swap]{d}{f_I}
      &K \arrow{d}{n} \\
      I \arrow[swap]{r}{m} & J
    \end{tikzcd}
  \]
  The commuting condition for cones states however that
  $f_J = m\circ f_I$, and $f_J = n\circ f_K$. Since $f_J$ is uniquely determined
  by composition of other arrows, we do not need to draw it and leave it implicit.
  With this in mind a limit over this diagram looks like this:
  \[
    \begin{tikzcd}
      {\color{red}V}
      \arrow[bend left=50,color=red]{drr}{g_K}
      \arrow[bend right=50,color=red,swap]{ddr}{g_I}
      \arrow[dashed,swap]{dr}{h}
      && \\
      &{\color{blue}U}
      \arrow[color=blue]{r}{f_K}
      \arrow[color=blue, swap]{d}{f_I}
      &K \arrow{d}{n} \\
      & I \arrow[swap]{r}{m} & J
    \end{tikzcd}
  \]
  The commuting condition for limits results in exactly the known commuting condition for
  pullbacks. We have therefore verified that pullbacks are a valid instance of the general limit definition.
  It follows similarly that equalizers are limit instances over diagrams of the shape
  \begin{tikzcd}
    I \arrow[r,shift left, "m"] \arrow[r,shift right,swap,"n"] & J
  \end{tikzcd}.
\end{example}

\chapter{Building new limits}
It is easy to see that limits do not necessarily exists in all categories.
Sometimes we want to show that a category has certain limits,
but it might be difficult to show for this category that the limits
of interest exist.
We now study how some limits can be used to construct other limits.
We start with an example.

\begin{example}
  Let $\mathscr C$ be a category with all products and all equalizers,
  that is, for all diagrams of the respective shape, there exists the product
  and the equalizer.
  We can use these limits to construct arbitrary pullbacks.
  We start with the diagram for pullbacks.
  \[
    \begin{tikzcd}
      & K \arrow{d}{n} \\
      I \arrow[swap]{r}{m} & J
    \end{tikzcd}
  \]
  Now we take the product of $I$ and $K$. This product exists by assumption.
  \[
    \begin{tikzcd}
      I \times K \arrow{r}{p_K} \arrow{d}{p_I} & K \arrow{d}{n} \\
      I \arrow[swap]{r}{m} & J
    \end{tikzcd}
  \]
  This looks already similar to a pullback.
  However, we need to ensure the commuting condition.
  For this purpose we use an equalizer which also exists by assumption.
  \[
    \begin{tikzcd}
      E \arrow{r}{e} &
      I \times K
      \arrow[r,shift right,swap, "m \circ p_I"] \arrow[r,shift left, "n \circ p_K"]
      & J \\
    \end{tikzcd}
  \]
  We now unfold this drawing.
  \[
    \begin{tikzcd}
      E \arrow[swap]{d}{p_I \circ e} \arrow{r}{p_K \circ e} & K \arrow{d}{n} \\
      I \arrow[swap]{r}{m} & J
    \end{tikzcd}
  \]
  $E$ satisfies the commuting and universality conditions of limits
  since it is already an equalizer.
  Therefore it is a pullback.
\end{example}

We now generalize
the approach used in the example. It turns out that we can construct all
limits with it.
Let $D : \bm I \to \mathscr C$ be any diagram.
We use the following construction:
\[
  \begin{tikzcd}
    E \arrow{r}{e} &
    \displaystyle \prod_{I \in \bm{I}} D(I)
    \arrow[r,shift left, "s"] \arrow[r,shift right,swap,"t"] &
    \displaystyle \prod_{K \overset{n}{\to} J \in \bm{I}} D(J)
  \end{tikzcd}
\]
with
\begin{align*}
  s_n &= Dn \circ p_K\ \hat{=} \prod_{I \in \bm{I}} D(I)
        \overset{p_K}{\rightarrow} D_K \overset{Dn}{\rightarrow} D_J \\
  t_n &= p_J\ \hat{=} \prod_{I \in \bm{I}} D(I)
        \overset{p_J}{\rightarrow} D_J
\end{align*}
The second product has components $p_n$ for each arrow
$K \overset{n}{\to} J \in \bm{I}$ with a projection on
the destination of the arrow.
For example for the diagram of a pullback
\[
  \begin{tikzcd}
    & K \arrow{d}{n} \\
    I \arrow[swap]{r}{m} & J
  \end{tikzcd}
\]
the mentioned product has projections $p_{1_I} = I$, $p_{1_J} = J$, and $p_{1_K} = K$
for each identity arrow.
Additionally there are projections for the remaining arrows $p_n = J$, and $p_m = J$.
Since $E$ is an equalizer we get
\begin{align*}
  s \circ e &= t \circ e\\
  s_n \circ e &= t_n \circ e, \quad \forall\ K \overset{n}{\rightarrow} J \in \bm I \\
  Dn \circ p_K \circ e &= p_J \circ e \\
  Dn \circ \uprho_K &= \uprho_J
\end{align*}
For all maps $K \overset{n}{\rightarrow} J$ we get the commuting diagram
\[
  \begin{tikzcd}
    E \arrow{r}{\uprho_K} \arrow{dr}[swap]{\uprho_J} & D_K \arrow{d}{Dn}\\
    & D_J
  \end{tikzcd}
\]
This is the commuting condition for a cone.
Since any other cone over $D$
is also a fork of $s$ and $t$
it follows that the equalizer $E$ is also a universal cone
and therefore a limit over $D$.

\begin{remark}
  We can even construct arbitrary limits over infinite diagrams
  because we assumed that $\mathscr C$ has all products.
  If $\mathscr C$ only has finite products, then we can
  use this construction for
  limits over finite diagrams only which is still useful since
  most practical limits are constructed using very small shapes for the diagram.\\
  Finite products themselves can in turn be constructed if $\mathscr C$
  has a terminal object and binary products. The terminal object is
  only needed for unary products. All other finite products
  can be constructed by cascading binary products.
\end{remark}

%% TODO: possibly other construction mentioned in awodey.
%% pullbacks + term constructs fin products + eq, which is the other direction
%% than the example.

\chapter{Colimits}
To define colimits we dualize the concept of limits.
\begin{definition}
  Let $D : \bm I \to \mathscr C$ be a diagram.
  A \textbf{cocone} $U$ over $D$ is an object $U \in \mathscr{C}$
  with maps $D_I \overset{ f_I}{\rightarrow} U,\ \forall\ I \in \bm{I}$, such that
  $\forall\ I \overset{m}{\to} J \in \bm{I}$, the following triangle commutes:
  \[
    \begin{tikzcd}
      D_I \arrow[rr, "D(m)"]\arrow[dr, swap, "f_I"] & & D_J \arrow{dl}{f_J}\\
      & U  &
    \end{tikzcd}
  \]
  That is $f_I = f_J \circ D(m)$.\\
\end{definition}
A cone has maps from the vertex into the diagram,
while a cocone has maps from the diagram into the vertex.
\begin{definition}
  A cocone $(D(I)\overset{f_I}{\to} U)_{I\in \bm{I}}$ is \textbf{universal}, if
  for all other cones $(\D(I) \overset{g_I}{\to} V)_{I\in \bm{I}}$ there is a unique
  map $h \from U \to V$, such that for all $I \in \bm{I}$
  the following triangles commute.
  \[
    \begin{tikzcd}
      D_I \arrow[dr, swap, "g_I"] \arrow{r}{f_I} & U \arrow{d}{h} \\
      & V
    \end{tikzcd}
  \]
  That is $g_i = f_i \circ h$.
\end{definition}

\begin{definition}
  A \textbf{colimit} is a universal co-cone over a diagram.
\end{definition}
%% TODO: complete definition
